---
title: "R Take-Home Assignment 5: Statistics with Penguins"
author: "Your Name"
date: "`r format(Sys.Date())`"
output:
  html_document:
    toc: true
    toc_depth: 2
    number_sections: true
  pdf_document:
    latex_engine: xelatex
---

## Overview

In this assignment you will practice simple statistical tests in R using the **palmerpenguins** dataset. We will ask biological questions about penguin body mass, learn to run standard tests in R, and interpret the results. Each section starts with a short story to explain why we are doing this step.

### Source

`palmerpenguins` package

### Skills

`t-tests, ANOVA, post-hoc tests, effect sizes, diagnostics`

```{r setup-ass5, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

## Preparation

You will need: `dplyr`, `ggplot2`, `palmerpenguins`, `broom`, `effectsize`, `tidyr`.

```{r}
## Task: load packages
## library(dplyr)
## library(ggplot2)
## library(palmerpenguins)
## library(broom)
## library(effectsize)
## library(tidyr)
```

## 1. Load the dataset

Before we can analyze penguins, we need to load the dataset. Real-world data often contain missing values, which can cause errors in analysis, so we will remove them first.

**Task:** Load the `penguins` dataset and remove missing values.

```{r}
## Task: load penguins dataset (hint: data(penguins))
## Task: remove rows with missing values (hint: drop_na())
```

## 2. Two-sample t-test

Imagine you are a biologist asking: *Do male and female penguins of the same species differ in body mass?*  
The **t-test** compares the means of two groups. It answers whether the difference is large enough that it is unlikely to be due to chance.

**Task:** Filter the dataset to one species (e.g. Adelie) and compare male vs female body mass with a t-test.

```{r}
## Task: filter to one species (hint: filter(species == "Adelie"))
## Task: run t.test(body_mass_g ~ sex, data = ...)
```

## 3. ANOVA

Now suppose we want to compare *all three species at once*. Do Gentoo, Adelie, and Chinstrap penguins differ in body mass?  
An **ANOVA (Analysis of Variance)** extends the t-test to more than two groups.

**Task:** Run an ANOVA with species as the predictor.

```{r}
## Task: run aov(body_mass_g ~ species, data = penguins)
## Task: check summary()
```

## 4. Post-hoc tests

ANOVA tells us if there is *some* difference between groups, but not *which groups differ*. For that, we run a **Tukey HSD post-hoc test**.

**Task:** Use `TukeyHSD()` on your ANOVA model to see which pairs of species differ.

```{r}
## Task: run TukeyHSD() on your ANOVA model
```

## 5. Effect sizes

Statistical significance is not enough: we also want to know *how large the difference is*.  
- For t-tests, we use **Cohen’s d** (small ≈ 0.2, medium ≈ 0.5, large ≈ 0.8).  
- For ANOVA, we use **η² (eta squared)** (small ≈ .01, medium ≈ .06, large ≈ .14).  

**Task:** Calculate Cohen’s d for your t-test and η² for your ANOVA.

```{r}
## Task: use effectsize::cohens_d() and effectsize::eta_squared()
```

## 6. Diagnostics (checking assumptions)

Every statistical test makes assumptions. For ANOVA, the two most important are:

1. **Equal spread of errors (homoscedasticity):**  
   The variation in the data should be roughly the same across all groups.  

2. **Normally distributed errors:**  
   The differences between the observed data and the model’s predictions (residuals) should follow a bell-shaped curve.  

We can check these assumptions using two standard diagnostic plots:

- **Residuals vs Fitted plot**  
  - Each dot is one observation.  
  - The x-axis shows what the model predicts, the y-axis shows the “error” (residual).  
  - What to look for: the dots should look like random noise.  
    - If you see a curve → the relationship might not be captured well.  
    - If you see a funnel shape → group variances may not be equal.  

- **Normal QQ plot**  
  - This compares your residuals to what would be expected if they were perfectly normal.  
  - What to look for: the dots should fall roughly along the diagonal line.  
    - If they bend away strongly → residuals may not be normally distributed.  

**Task:** Create both plots for your ANOVA model and then write 2–3 sentences interpreting them.

```{r}
## Task: residual vs fitted
## Task: QQ plot
```

## 7. Reflection

Statistics are not just numbers: they help answer real questions. Think like a biologist preparing a short report.

**Task:** Write 3–4 sentences: what did you learn about differences in penguin body mass across species and sexes? How strong are the effects?
